Untitled

### 2022-04-25 Papers/Polymethylation_scores/04_background.Rmd
Betsy edits

In 2018, 11% of pregnant people in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Children exposed to cigarette smoking *in utero* are more likely to have low-birth weight, negative neurodevelopmental outcomes, and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior can be challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard smoking biomarker, serum cotinine levels, has a half-life of nine hours in pregnant women, and is rarely available outside of birth cohort research [@Dempsey:2002wr]. Alternative biomarkers of prenatal maternal smoke exposure with a larger window of measurement could improve research and clinical practice.

DNA methylation is a possible biomarker to capture prenatal smoking exposure. Prenatal maternal smoking is associated with offspring DNA methylation signatures at birth in placental and cord blood samples, and postnatally in peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb]. Another postnatal biospecimen option is saliva, which is easier to collect than blood and increasingly used in large epidemiological cohorts [@Langie2017]. Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues [@Bakulski2016]. In addition, most studies of prenatal maternal smoking and childhood DNA methylation have been conducted in primarily European genetic ancestry populations [@Joubert:2016ue; @Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty], and it will be important to test generalizability to other groups. Few studies have assessed multiple genetic ancestry groups in the same cohort and no study has evaluated associations between maternal pregnancy cigarette smoking and salivary DNA methylation in children.

>>>>

Children exposed to cigarette smoking *in utero* are more likely to have low-birth weight, negative neurodevelopmental outcomes, and asthma [@Banderali:2015vu]. Although 11% of pregnant people in the United States in 2018 reported any past-month cigarette use [@SAMHSA2018], this is considered an underestimate. When compared to biomarker measures, maternal prenatal smoking is underreported - presumably due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard smoking biomarker, serum cotinine levels, has a half-life of nine hours in pregnant women, and is rarely available outside of birth cohort research [@Dempsey:2002wr]. Alternative biomarkers of prenatal maternal smoke exposure with a larger window of measurement could improve measurement of prenatal smoke exposure in older children and adults.

DNA methylation is a possible biomarker of prenatal smoke exposure. Prenatal maternal smoking is associated with offspring DNA methylation signatures at birth in placental and cord blood samples, and postnatally in peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb]. Another postnatal biospecimen option is saliva, which is easier to collect than blood and increasingly collected in large epidemiological cohorts [@Langie2017]. However, associations of prenatal maternal smoking with salivary DNA methylation have not been measured. Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues [@Bakulski2016]. In addition, most studies of prenatal maternal smoking and childhood DNA methylation have been conducted in primarily European genetic ancestry populations [@Joubert:2016ue; @Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty], and it will be important to test generalizability to other groups. Few studies have assessed multiple genetic ancestry groups in the same cohort and no study has evaluated associations between maternal pregnancy cigarette smoking and salivary DNA methylation in children.

### 2022-04-25 Papers/Polymethylation_scores/05_results.Rmd
Betsy edits

In a sample of the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r m_og_methyl` samples from `r n_og_methyl` unique child participants with the Illumina 450K array. Complete data on covariates of interest was available on `r m_obs` samples from `r n_total` participants who were included in the analysis (Figure 1). There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples. Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African genetic ancestry and less likely to be from children of Admixed Latin genetic ancestry (Supplemental Table 1). In the included sample, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use. The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`. Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Admixed ancestry - Latin heritage']`% were of Admixed Latin genetic ancestry.

 All of the polymethylation scores for prenatal smoke exposure were correlated (Supplemental Figure 1). Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $r$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).

@@

In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* DNA methylation sites identified from previous meta-analyses. We replicated the association of prenatal maternal smoking with these DNA methylation at GFI1: cg14179389, CYP1A1: cg05549655, MYO1G: cg22132788, and MYO1G: cg04180046  (Supplemental Figure 8).

@@

Next, we assessed whether including DNA methylation summary measures increased the accuracy of classifying prenatal smoke exposure over a base model consisting of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing, the first two genetic principal components and an indicator variable for oversampled city. At age 9, including the polymethylation score increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points over the nested base model (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7) %>%format(., scientific=F)`). At age 15 including the polymethylation score for prenatal smoke exposure increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)` percentage points (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7)%>%format(., scientific=F)`). Including DNA methylation at *AHRR*: cg0557591 with base model covariates did not improve the classification compared to using base model covariates alone  (Supplemental Tables 4 and 5).  Classification accuracy when including the polymethylation score from btoh age 9 and 15 was very similar to classification accuracy when using the polymethylation score from age 9 only (Figure 4B).

>>>>

Among selected participants in the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r m_og_methyl` samples from `r n_og_methyl` unique child participants with the Illumina 450K array. Complete data on covariates of interest was available on `r m_obs` samples from `r n_total` participants who were included in the analysis (Figure 1). There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples. Excluded participants were similar to included participants, except that included participants were slightly more likely to be of African genetic ancestry and less likely to be of Admixed Latin genetic ancestry (Supplemental Table 1). Among included participants, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use. The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`. Of the included children, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Admixed ancestry - Latin heritage']`% were of Admixed Latin genetic ancestry.

 All of the polymethylation scores for prenatal smoke exposure were correlated with each other (Supplemental Figure 1). Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $r$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).

@@

In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* DNA methylation sites identified from previous meta-analyses. We replicated the association of prenatal maternal smoking with DNA methylation at GFI1: cg14179389, CYP1A1: cg05549655, MYO1G: cg22132788, and MYO1G: cg04180046  (Supplemental Figure 8).

@@

Next, we assessed whether including DNA methylation summary measures increased the accuracy of classifying prenatal smoke exposure over a base model consisting of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing, the first two genetic principal components and an indicator variable for oversampled city. At age 9, including the polymethylation score increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points over the nested base model (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7) %>%format(., scientific=F)`). At age 15 including the polymethylation score for prenatal smoke exposure increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)` percentage points (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7)%>%format(., scientific=F)`). Including DNA methylation at *AHRR*: cg0557591 with base model covariates did not improve the classification compared to using base model covariates alone  (Supplemental Tables 4 and 5).  Classification accuracy when including the polymethylation score from both age 9 and 15 was very similar to classification accuracy when using the polymethylation score from age 9 only (Figure 4B).

### 2022-04-25 Papers/Polymethylation_scores/08_methods.Rmd
Betsy edits

The Fragile Families and Child Wellbeing Study is a birth cohort of nearly 5,000 children born in 20 cities in the United States between 1998 and 2000 [@Reichman:2001wv]. Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv]. Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, the mother or baby were too ill to complete the interview, or the baby died before the interview could take place. Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; an age 22 follow up is ongoing. Across certain child ages, assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child. At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv]. A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing [@Reichman:2001wv]. To be eligible for the DNA methylation assessment children had to have participated and given saliva at age 9 and 15; after which a random sample of these eligible participants was then selected. One exception to this rule is that all participants in the Study of Adolescent Neurodevelopment (SAND) were assayed even if they only provided a sample at one time-point. All analyses account for oversampling of these participants.

@@

Prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth). At baseline, mothers answered categorical questions about their prenatal smoking, drug and alcohol use. For maternal prenatal smoking, mother's were asked: "During your pregnancy, how many cigarettes did you smoke?". Responses were categorical: 2 or more packs a day, 1 or more but less than 2 packs per day, less than 1 pack a day or none. Few participants reported smoking a pack or more a day, thus for our primary exposure variable we created a dichotomous any versus no prenatal smoke exposure variable.

In sensitivity analyses we controlled for other types of prenatal exposures. Mothers were also asked how frequently they drank during their pregnancy. We created a dichotomous any versus no prenatal alcohol exposure variable. Similarly, mothers were asked how frequently they used any other drug during their pregnancy. We again created a binary variable for any versus no prenatal use of other drugs.

In sensitivity analyses, we controlled for postnatal second-hand smoke exposure. When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about smoking in the household. To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for postnatal exposure at ages 1 and/or 5. To encapsulate recent postnatal smoke exposure to the age 9 and 15 interview where DNA methylation was collected, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to each interview.

To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking. At age 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no), and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).

@@

Our primary outcome variable was a polymethylation score for prenatal maternal smoke exposure. From an independent meta-analysis of prenatal smoke exposure and DNA methylation (the Joubert meta-analysis), we extracted the regression coefficients of 6,074 DNA methylation sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue]. We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum. For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control.

To assess the robustness of polymethylation scores to the source of coefficients for weights, we generated two sensitivity polymethylation scores from the Joubert meta-analysis [@Joubert:2016ue]. First, we used the regression of 568 cord blood DNA methylation sites associated with prenatal maternal smoking at a strict Bonferroni threshold, when analyses were not adjusted for cell proportions [@Joubert:2016ue]. Second, we used the regression coefficients of 19 child peripheral blood DNA methylation sites associated with prenatal maternal smoking at the strict Bonferroni threshold when analyses were not adjusted for cell proportions [@Joubert:2016ue].

@@

We excluded samples missing any data on the following variables: maternal prenatal smoking, maternal prenatal alcohol use, maternal prenatal other drug use, maternal income to poverty ratio, postnatal secondhand smoke exposure variables, child sex, or child age. We also excluded samples that were missing genetic data. Age 9 samples from children who reported ever smoking a cigarette or using tobacco at age 9 or were missing own-smoking data were excluded. Age 15 samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded. Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample.

>>>>

The Fragile Families and Child Wellbeing Study is a birth cohort of nearly 5,000 children born in 20 cities in the United States between 1998 and 2000 [@Reichman:2001wv]. Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv]. Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, the mother or baby were too ill to complete the interview, or the baby died before the interview could take place. Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; at age 22 follow up is ongoing. Across certain child ages, assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child. At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv]. A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing [@Reichman:2001wv]. To be eligible for the DNA methylation assessment children had to have participated and given saliva at age 9 and 15; after which a random sample of these eligible participants was then selected. One exception to this rule is that all participants in the Study of Adolescent Neurodevelopment (SAND) were assayed even if they only provided a sample at one time-point. All analyses account for oversampling of these participants.

@@

At time of child's birth, mothers answered categorical questions about their prenatal smoking, drug and alcohol use. For maternal prenatal smoking, mothers were asked: "During your pregnancy, how many cigarettes did you smoke?". Responses were categorical: 2 or more packs a day, 1 or more but less than 2 packs per day, less than 1 pack a day or none. Few participants reported smoking a pack or more a day, thus for our primary exposure variable we created a dichotomous any versus no prenatal smoke exposure variable.

Mothers were also asked how frequently they drank during their pregnancy. We created a dichotomous any versus no prenatal alcohol exposure variable. Similarly, mothers were asked how frequently they used any other drug during their pregnancy. We again created a binary variable for any versus no prenatal use of other drugs.

When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about smoking in the household. To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for postnatal exposure at ages 1 and/or 5. To encapsulate recent postnatal smoke exposure to the age 9 and 15 interview where DNA methylation was collected, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to each interview.

At age 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no), and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).

@@

Our primary outcome variable was a polymethylation score for prenatal maternal smoke exposure. From an independent meta-analysis of prenatal smoke exposure and DNA methylation (the 2016 Joubert meta-analysis), we extracted the regression coefficients of 6,074 DNA methylation sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue]. We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum. For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control.

To assess the robustness of polymethylation scores to the source of coefficients for weights, we generated two sensitivity polymethylation scores from the 2016 Joubert meta-analysis [@Joubert:2016ue]. First, we used the regression of 568 cord blood DNA methylation sites associated with prenatal maternal smoking at a strict Bonferroni threshold, when analyses were not adjusted for cell proportions [@Joubert:2016ue]. Second, we used the regression coefficients of 19 child peripheral blood DNA methylation sites associated with prenatal maternal smoking at the strict Bonferroni threshold when analyses were not adjusted for cell proportions [@Joubert:2016ue].

@@

We excluded samples missing any data on the following variables: maternal prenatal smoking, maternal prenatal alcohol use, maternal prenatal other drug use, maternal income to poverty ratio, postnatal secondhand smoke exposure variables, child sex, or child age. We also excluded samples that were missing genetic data. To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking. Age 9 samples from children who reported ever smoking a cigarette or using tobacco at age 9 or were missing own-smoking data were excluded. Age 15 samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded. Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample (Figure 1).
